A coherent and comprehensive dating system for stars in the Milky Way

### 2017-02-06 documents/ms.tex
adding words on the subway

\newcommand{\columbia}{1}
\newcommand{\simons}{2}
\newcommand{\nyu}{3}
\newcommand{\cca}{4}
\newcommand{\mpia}{5}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\begin{abstract}
Words go here.
\end{abstract}

@@

The processes behind the formation of the Milky Way and the objects within it
are some of the most elusive and complicated topics in astronomy today,
connected by a common theme: stellar ages.
Ages provide the key to understanding the evolution of all astrophysical
objects, but age, unfortunately, one of the most difficult stellar properties
to measure.
Different dating methods often produce inconsistent predictions for the age of
the same star.
This is in part because the underlying processes generating the evolution of
the observable properties are different and in part because our understanding
of the underlying physics is flawed or incomplete.
In table \ref{tab:dating_methods} we present a list of dating methods, the
main observables associated with that method, the underlying physics driving
the changing observables, the types of star the method applies to and the
empirical or physical nature of the model.

@@

@@

\section{Method}

\section{Results and Discussion}

@@

% acknowledgements
This research was funded by the Simons Foundation.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
Kepler mission is provided by the NASA Science Mission directorate.

>>>>

\newcommand{\racomment}[1]{{\color{red}#1}}

% Affiliations
\newcommand{\columbia}{1}
\newcommand{\simons}{2}
\newcommand{\nyu}{3}
\newcommand{\cca}{4}
\newcommand{\mpia}{5}

% Missions
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\tess}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\lsst}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\sdss}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\plato}{{\it PLATO}}
\newcommand{\TGAS}{{\it TGAS}}
\newcommand{\tgas}{{\it TGAS}}
\newcommand{\APOGEE}{{\it APOGEE}}
\newcommand{\apogee}{{\it APOGEE}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\Gaia}{{\it Gaia}}

% Stellar parameters
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

% Common expressions
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\etal}{{\it et al.}}

% Paper-specific
\newcommand{\nkictags}{2000}

@@

\begin{abstract}
% Aims
Stellar age underpins a wide range of astronomical applications, yet
remains one of the most challenging properties to infer.
% Method
Competing dating methods with different observational inputs often produce
    wildly inconsistent age estimates.
In this work we attempt to combine multiple methods into one coherent model
    and calibrate that model using galactic dynamics --- a well understood
    physical process.
We identify a sample of stars with at least two of the following dating
    methods available: rotation-dating, isochronal-dating,
    asteroseismic-dating and dynamical-dating.
% Results
\end{abstract}

@@

The processes behind the formation of the Milky Way and the objects within it
are some of the most elusive and complicated topics in astronomy today,
connected by a common theme: stellar ages.
Ages provide the key to understanding the evolution of all astrophysical
objects, but age is, unfortunately, one of the most difficult stellar
properties to measure.
Different dating methods often produce inconsistent predictions for the age of
the same star.
This is in part because the underlying processes generating the evolution of
the observable properties are different and in part because our understanding
of the underlying physics is flawed or incomplete.
The various available dating methods can be categorised by the underlying
physical process they trace --- the radial extent of the hydrogen-burning core
or the evolving magnetic dynamo for example.
In addition, they can be classified by their level of empiricism, \ie\ the
number of free parameters that need to tuned when calibrating these models.
The physics behind orbital heating, for example, is very well understood and
has only one tunable parameter.
On the other hand, magnetic activity evolution is poorly understood and the
dating methods that are based on it are often entirely empirical.
In table \ref{tab:dating_methods} we provide an overview of various dating
methods, the main observables associated with them, the underlying physics
driving the changing observables, the types of star the method applies to and
the empirical or physical nature of the model.

@@

In this paper we focus on four of these dating methods: rotation,
asteroseismology, isochrones and galactic dynamics.
A brief overview of each of these methods is provided in below.

@@

Not only do a number of dating methods exists, several different models are
often available for the same dating method.
In the case of rotation-dating...

\racomment{What happens when you can't measure a rotation period?}
\racomment{What about the difference in physics that you are probing?}

\subsection{Asteroseismology}

\subsection{Isochrones}

\subsection{Galactic Dynamics}

This paper is laid out as follows.
% In \textsection \ref{sec:dating_methods} we provide an overview of each dating
% method used in this paper and demonstrate their inconsistency in age
% predictions.
In \textsection \ref{sec:data} we describe the data used to calibrate our new
model, which is described in \textsection \ref{sec:method}.
In \textsection \ref{sec:results} we present a new catalogue of ages.

\section{Data}
\label{sec:data}
We assembled a sample of stars that have at least two of the four dating
methods available.
Our main asteroseismic sample targets comprise the $~$500 short cadence
\kepler\ prime targets.
These stars all have light curves and therefore rotation periods (or potential
rotation periods).
The majority are also in Gaia DR1, therefore have parallaxes, positions and
2-dimensional velocities, which can provide constraints on their vertical
actions.
A subset of these stars have spectroscopic information and therefore make
excellent isochronal targets.
Our main rotation sample is the \nkictgas\ stars in both the \kepler\ input
catalogue (KIC) and \TGAS.
These stars have potential rotation periods, parallaxes, positions and
2-dimensional velocities.
Our main dynamics data set is the intersection of the \apogee\ and \tgas\
catalogues.
The majority of these stars have ages from {\tt the cannon} \cite{Ness2015},
plus \gaia\ parallaxes, positions and proper motions as well as radial
velocities, providing vertical actions calculated from the full 3-dimensional
phase-space.
They are mostly red giants.

\section{Method}
\label{sec:method}

In order to establish the level of inconsistency between dating methods, we
first compare predictions for each of the above data sets.
This is purely an illustrative test as these dating methods are not always
independent, for example many rotation-dating models are calibrated using
asteroseismic age predictions.
For the asteroseismic sample we find that

\section{Results and Discussion}
\label{sec:results}

@@

% acknowledgements
This research was funded by the Simons Foundation.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
Kepler mission is provided by the NASA Science Mission directorate.
\racomment{Add TGAS \& Apogee acknowledgements.}
